### 2025-05-05 sec/4_evaluation.tex
Conc

\section{Experimental evaluation}
\label{sec:evaluation}
In this section we evaluate our Branch and Bound solver that uses our novel convex relaxation on point cloud registration and related outlier robust problems. 
We measure the required number of iterations as well as the per-iteration costs. For this we have implemented a simple version of a best-first Branch-and-Bound as in \cite{10.1007/978-3-642-37444-9_42}.

We also have validated the corecteness of the algorithm and it's implementation empirically: We did not find any instance in which it did not find the global minimum of the TLS problem.

@@

\subsection{Synthetic experiments}
\begin{figure}[!ht]
	\centering
	\includegraphics[width=0.49\linewidth]{figures/execution_time_over_outlier_rate_boxplot}
	\includegraphics[width=0.49\linewidth]{figures/rotation_error_over_outlier_rate.pdf}
	\caption{Execution time and rotation error over different outlier rates with $N=100$ on synthetic data.}
	\label{fig:execution-time-over-outlier-rate}
\end{figure}

\begin{figure}[!ht]
	\centering
	\begin{adjustbox}{width=1.\linewidth}
		\includegraphics{figures/robust_wahba_runtime_rotation_error.pdf}
	\end{adjustbox}
	\caption{Rotation-only experiments: The average rotation error over the ourlier rate: The Truncated Least Squares method is robust up to 95\% outliers.}
	\label{fig:rotation-error-over-outlier-rate}
\end{figure}

\begin{figure}[!ht]
	\centering
	\begin{adjustbox}{width=1.\linewidth}
		\includegraphics{figures/robust_wahba_runtime_execution_time.pdf}
	\end{adjustbox}
	\caption{Average runtime of our Branch-and-Bound solver depending on the total amount of point correspondences $N$.}
	\label{fig:runtime}
\end{figure}

@@

\subsubsection{Scalability}
\begin{figure}[!ht]
	\centering
	\includegraphics[width=0.9\linewidth]{figures/scalability-synth}
	\caption{Scalability on synthetic data, outlier rate $75\%$}
	\label{fig:execution-time-over-n}
\end{figure}

@@

\subsubsection{Validity of the global optimality certificate}

We test on synthetic data whether the global optimality certificate is valid, i.e. whether at termination, the lower bound is less or equal than the TLS objective at the ground-truth solution.
We define the suboptimality $\eta$ of the solution based on the upper bound (UB), that is, the best known solution, and the lower bound (LB) \cite{9785843}
\begin{equation}
	\begin{aligned}    
		\eta = \frac{\text{UB} - \text{LB}}{1 + |\text{UB}| + |\text{LB}|}
	\end{aligned}
\end{equation}
We test with $N=100$, $\epsilon=0.5$ and vary the data scale $s$ from 3 to 100. We test for each scale outlier rates of $0\%, 30\%, 50\%, 90\%, 96\%$, and do 1000 trials for each combination.
\begin{figure}[!ht]
	\centering
	\includegraphics[width=0.49\linewidth]{figures/certificate_validity_vs_scale}
	\includegraphics[width=0.49\linewidth]{figures/eta_subopt_over_outlier_rate}
	\caption{Amount of instances with valid global optimality certificate (i.e. valid suboptimality) on synthetic data, including achieved $\eta$-suboptimality.}
	\label{fig:cert-validity}
\end{figure}

@@

\begin{figure}[!ht]
	\centering
	\includegraphics[width=0.49\linewidth]{figures/subopt_stride}
	\includegraphics[width=0.49\linewidth]{figures/runtime_stride}
	\caption{The reached suboptimality and the runtime using STRIDE over the outlier rate on synthetic data, $N=30$.}
	\label{fig:stride-subopt-eval}
\end{figure}
\begin{figure}[!ht]
	\centering
	\includegraphics[width=0.9\linewidth]{figures/rotation_error_stride}
	\caption{Rotation error using STRIDE.}
	\label{fig:stride-rot-err}
\end{figure}

The results are shown in Fig. \ref{fig:stride-subopt-eval}, STRIDE is able to find and certify the global optimum as long as the outlier rate is below 90\%. The runtime of STRIDE is even with as few as 30 points in the order of $10^3$ seconds. STRIDE achieves a similar rotation error compared to our solver in instances that could be solved (Fig. \ref{fig:stride-rot-err}). Note that the rotation error is meaningless for instances that STRIDE could not solve (i.e. above 80\%), since it is always able to detect them.

\subsection{Results on 3DMatch dataset}
\input{sec/eval-3DMatch-table}
We evaluate our solver on the 3DMatch dataset \cite{zeng20163dmatch} 
following \cite{SC2-PCR-Chen-2022-CVPR, zhang20233d} on all scenes. We evaluate the residual rotation (RE) and translation (TE) and the registration recall (RR). A registration is defined as successful if RE $\leq 15$ degrees and TE $\leq 30$ cm. We sub-sample the correspondeces to 500 and set the inlier-threshold to $\epsilon = 0.6$. For the other methods, we report the results from \cite{zhang20233d} since the test setup is identical. The results are shown in Tab. \ref{tab:3DMatch-results}.

>>>>

\section{Experimental evaluation}
\label{sec:evaluation}
In this section we evaluate a Branch and Bound solver that uses our novel convex relaxation on point cloud registration and related outlier robust problems. We have implemented a basic version of BnB using best-first search. To branch over the spherical space of 3D rotations, we use rotation vector approach as in \cite{10.1007/978-3-642-37444-9_42}.

We also have validated the correctness of the algorithm and it's implementation empirically: We did not find any instance in which the lower bound was invalid, i.e. higher than the ground-truth global minimum.

@@

\subsection{Translation experiments}
We evaluate solving the translation and the rotation problem separately.

\subsection{Synthetic experiments}
\begin{figure}[!ht]
	\centering
%	\includegraphics[width=0.49\linewidth]{figures/execution_time_over_outlier_rate_boxplot}
%	\includegraphics[width=0.49\linewidth]{figures/rotation_error_over_outlier_rate.pdf}
	\caption{Execution time and rotation error over different outlier rates with $N=100$ on synthetic data.}
	\label{fig:execution-time-over-outlier-rate}
\end{figure}

\begin{figure}[!ht]
	\centering
	\caption{Rotation-only experiments: The average rotation error over the outlier rate: The Truncated Least Squares method is robust up to 95\% outliers.}
	\label{fig:rotation-error-over-outlier-rate}
\end{figure}

\begin{figure}[!ht]
	\centering
	\caption{Average runtime of our Branch-and-Bound solver depending on the total amount of point correspondences $N$.}
	\label{fig:runtime}
\end{figure}

@@

\subsubsection{Scalability}
\begin{figure}[!ht]
	\centering
	%\includegraphics[width=0.9\linewidth]{figures/scalability-synth}
	\caption{Scalability on synthetic data, outlier rate $75\%$}
	\label{fig:execution-time-over-n}
\end{figure}

@@

\subsection{Rotation experiments}
We solve the rotation-only truncated least squares problem with the same approach as the translation problem: We use the same BnB solver and convex relaxation, only the computation of the residual intervals changes, as well as the minimization of the relaxation function.

\subsubsection{Validity of the global optimality certificate}
We test on synthetic data whether the global optimality certificate is valid, i.e. whether at termination, the lower bound is less or equal than the TLS objective at the ground-truth solution.
We define the suboptimality $\eta$ of the solution based on the upper bound (UB), that is, the best known solution, and the lower bound (LB) \cite{9785843}
\begin{equation}
	\begin{aligned}    
		\eta = \frac{\text{UB} - \text{LB}}{1 + |\text{UB}| + |\text{LB}|}
	\end{aligned}
\end{equation}
We test with $N=100$, $\epsilon=0.5$ and vary the data scale $s$ from 3 to 100. We test for each scale variation the outlier rates $0\%, 30\%, 50\%, 90\%, 96\%$. We do 1000 trials for each such parameter set.

\begin{figure}[!ht]
	\centering
	%\includegraphics[width=0.49\linewidth]{figures/certificate_validity_vs_scale}
	%\includegraphics[width=0.49\linewidth]{figures/eta_subopt_over_outlier_rate}
	\caption{Amount of instances with valid global optimality certificate (i.e. valid suboptimality) on synthetic data, including achieved $\eta$-suboptimality.}
	\label{fig:cert-validity}
\end{figure}

@@

\begin{figure}[!ht]
	\centering
	%\includegraphics[width=0.49\linewidth]{figures/subopt_stride}
	%\includegraphics[width=0.49\linewidth]{figures/runtime_stride}
	\caption{The reached suboptimality and the runtime using STRIDE over the outlier rate on synthetic data, $N=30$.}
	\label{fig:stride-subopt-eval}
\end{figure}
\begin{figure}[!ht]
	\centering
	%\includegraphics[width=0.9\linewidth]{figures/rotation_error_stride}
	\caption{Rotation error using STRIDE.}
	\label{fig:stride-rot-err}
\end{figure}

The results are shown in Fig. \ref{fig:stride-subopt-eval}, STRIDE is able to find and certify the global optimum as long as the outlier rate is below 90\%. The runtime of STRIDE is even with as few as 30 points in the order of $10^3$ seconds. In instances that STRIDE could solve (i.e. where the relaxation is tight), it achieves a similar rotation error compared to our solver (Fig. \ref{fig:stride-rot-err}). Note that the rotation error is meaningless for instances that STRIDE could not solve (i.e. above 80\% outliers).

\subsection{Results on 3DMatch dataset}
\input{sec/eval-3DMatch-table}
We evaluated our solver on the 3DMatch dataset \cite{zeng20163dmatch} 
following the same protocol as in \cite{SC2-PCR-Chen-2022-CVPR, zhang20233d}. We use all scenes and evaluate the residual rotation error (RE) and translation error (TE) as well as the registration recall (RR). A registration is defined as successful if RE $\leq 15$ degrees and TE $\leq 30$ cm holds. We sub-sample the correspondences to 500 and set the inlier-threshold to $\epsilon = 0.6$. For the other methods, we report the results from \cite{zhang20233d} since the test setup is identical. The results are shown in Tab. \ref{tab:3DMatch-results}.

### 2025-05-05 sec/5_conclusion.tex
Conc

In this paper, we presented a general approach of minimizing the Truncated Least Squares (TLS) objective motivated by the geometry of the problem. We showed how intersection graphs allow for an elegant formulation of the combinatorial problem and how the global minimum can be found among the maximal cliques. We showed that surprisingly, the previously studied compatibility graphs motivated by problem-specific geometric observations lead to these special classes of graphs, namely the intersection graphs. 
We give a sufficient condition under which the global optimum is a maximal clique and discuss cases where subcliques may be the global minimum.
Unfortunately, we were not able to fully prove that we can detect all conditions under which subcliques may be the global optimum.
We think that a better understanding of the $\SEthree$-manifolds and their geometry including their combinatorial properties could lead to efficient algorithms for detecting such cases.

>>>>

In this paper, we presented a general approach of minimizing the Truncated Least Squares (TLS) objective function 
using a fast linear time convex relaxation and Branch-and Bound. Our convex relaxation is general and can be applied for every model such as translation, rotation and rigid transform as long as residual intervals can be computed. We gave a full proof
for the correctness of the relaxation, which is the main part of the proof of global optimality of a BnB algorithm.
Our evaluation shows that the cost to compute and minimize the relaxation is in the single milliseconds range, allowing to solve the 
translation and rotation-only problem two orders of magnitude faster than with the SoTA SDP-based method STRIDE.
More work is however needed to solve the full 6DoF rigid transformation problem. Altought our relaxation can be applied here as well, branching over the 6D space is overall too slow. Further work in this direction will focus on contracting the rotation search space given a translation space, so that the search is not excessively inefficient.
